Titeln på avhandlingens språk

### 2023-05-19 body/part1/chapter3-contributions/wca-implications.tex
reformulating

In this section, we aim to delve deeper into the implications of our methodology for the optimization of \gls{WCA}.
In particular, we wish to explore the impact improving the accuracy of the client-side emulation has on the potential for optimization of the applications under evaluation.
We approach this question by developing an improved, more detailed version of our model of human behavior in \gls{WCA} used in the EdgeDroid tool, which we then incorporate into our methodology along with a mathematical framework to study the potential for optimizing \gls{WCA} applications.

Our work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} has used a model that emulates a rigid and unyielding human who makes no mistakes and is not affected by low responsiveness in the system.
However, this is not an accurate representation of real human behavior, and a more realistic human model must take into consideration the perceived feedback latency and its effects on human reactions.
In order to develop a more accurate human model in the context of \gls{WCA}, we first conduct an experimental study that measures human performance in different scenarios, taking into account the impact of system latency on the behavior of humans.
The results of this study serve as the foundation for building a more representative model of human behavior.

Once we have established a more accurate model of human behavior, we can explore its implications for the optimization of \gls{WCA} through our methodology.
By improving the accuracy of our model of human behavior, we aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be effectively applied to address these.
This research highlights the importance of the accuracy of emulations as  crucial factor in the application of our methodology.

@@

In order to more easily and accurately discuss the implications of our methodology for \gls{WCA}, we first characterize these applications and provide some definitions relating to their operation.
We focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistance applications which have as their goal the guidance of a user through a task, using a series of instructions.

As discussed above, \glspl{WCA} track the state of the real world through sampling of inputs, most commonly of video feeds.
In step-based \glspl{WCA}, the state in question refers to the progress of some arbitrary, sequential task, such as the assembly of a LEGO model or of a piece of IKEA furniture~\cite{chen2018application}.
We will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided;
the time interval between these two events will be referred to as the \emph{step duration}.

@@

In the following, please refer to \cref{fig:wcastep}.
\ensuremath{\{ t_0, t_1, \ldots, t_{n + 1} \}} corresponds to a series of discrete and sequential sampling instants, such that \ensuremath{t_0} corresponds to the instant at which the instruction for step \ensuremath{M} is provided and the first sample of the step is taken.
\ensuremath{t_n} corresponds to the instant at which the final sample of step \ensuremath{M} is taken;
this sample is the first sample to capture the finished state of the step.
At this point, the application transitions to the next step \ensuremath{M + 1}, and thus \ensuremath{t_{n + 1}} is the instant at which the next instruction is provided and step \ensuremath{M + 1} begins.
We will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.

\glsreset{TTF}%
From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in \cref{sec:background:wca} and in \cref{fig:wca:operation}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
A corollary of this is that these are therefore also the \emph{only} points in time at which users can notice changes in system responsiveness.
If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
\begin{inlineenum}
    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
\end{inlineenum}
These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
We will refer to them extensively in \cref{chap:contributions}.

%\subsection{Mechanisms relating human behavior to system responsiveness}
%
%The question of how people respond to delay in a computer system is grounded in how people perceive time.
%Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
%More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
%The estimation of an event's duration varies with context of various types
%\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
%\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
%\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
%\item arousal can expand durations~\cite{droit_volet2011emotion}.
%\end{inlineenum}
%
%Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
%It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
%Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
%Violations of expectancy have a strong impact on the acceptability of computer systems.
%Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
%In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
%
%It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
%However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
%Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
%
%For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
%They then proposed a latency tolerance zone according to the task demands.
%For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
%Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
%
%While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
%These mechanisms could be cognitive or emotional in origin.
%
%A first possible explanation comes from research on cognitive and motor planning.
%Delay may move users from relatively automatic to more attention-demanding processing.
%Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
%As competency in a task increases, execution of the hierarchy becomes increasingly automated.
%Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
%Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
%
%Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
%Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
%
%A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
%As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
%
%Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
%
%The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
%If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
%Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
%Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
%Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
%
%Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
%Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
%High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
%Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
%Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting. Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
%These and other measures of individual variation were considered here.

\subsection{Characterizing human behavior}

\Cref{paper:olguinmunoz2021impact} aims to investigate the effects of system responsiveness on human behavior in step-based \gls{WCA}.
This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \gls{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.
The study involved \num{40} undergraduate students who interacted with a \gls{WCA} application while the system responsiveness was systematically altered in real-time, and their behavioral and physiological reactions were recorded.
The study attempted to answer a number of core research questions relating to human behavior in the context of these applications.
These included
\begin{inlineenum}
    \item whether users change their actions in relation to system latency
    \item whether they show signals of arousal in physiological responses
    \item whether their responses to delay effects are mediated by cognition and/or emotion
    \item whether these effects are mediated by personality indicators
\end{inlineenum}.

We found that reduced system responsiveness induces additional behavioral slow-down in \gls{WCA} users, and this effect scales with the strength of the reduction in system responsiveness.
Furthermore, we note that humans tend to speed up as the task progresses;
however, as system responsiveness decreases, this effect is strongly dampened.
The effects of reduced system responsiveness on users linger for at least four steps after the system responsiveness improves, leading to longer execution times for steps immediately after a period of reduced responsiveness.
Finally, the trait of \emph{neuroticism}~\cite{john1999big} seems to play a significant role in modulating users' response to reduced responsiveness in \gls{WCA}.

@@

We employed a version of the step-based \emph{Gabriel} cognitive assistant framework first introduced by \citeauthor{chen2018application}~\cite{chen2018application}.
The assistant was instrumented to capture key application and task performance metrics in real time, in particular step execution time.
Furthermore, the software was modified to allow for the real-time alteration of system responsiveness by extending the processing interval of each input to a temporarily fixed value denoted \emph{delay}.
This is illustrated in \cref{fig:regularwca-vs-delaywca}.
If, during a series of steps, delay was set to a value \ensuremath{\mathbb{D}}, the feedback for each input frame was provided to the user exactly a time interval \ensuremath{\mathbb{D}} after frame capture.
That is,
\begin{align}
    t_k - t_{k - 1} = \mathbb{D} \qquad \forall k \in [2, n + 1]
\end{align}
%Seven values were selected for \ensuremath{\mathbb{D}}: no added delay (which was referred to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second}.
In order to fully characterize the resulting \emph{\glspl{TTF}} perceived by participants, we note that the instant of final frame capture for a step, \ensuremath{t_c} (and conversely the wait time \( \mathcal{W} \) of the step), can be assumed to be uniformly distributed in \( [t_{n - 1}, t_n] \), without loss of generality.
Thus the average \gls{TTF} for a step can be expressed as
\begin{align*}
    TTF = (t_{n + 1} - t_{n}) + \frac{1}{2}(t_n - t_{n - 1})
\end{align*}
which, for a step with a given delay \ensuremath{\mathbb{D}}, works out to\footnote{%
    The \num{0}/no-delay case was handled by calculating the empirical average frame \gls{RTT} for these steps, and using this value as the delay \ensuremath{\mathbb{D}}.
    This yielded an average \gls{TTF} of \SI{0.57}{\second} for these steps.
}
\begin{align}
    TTF &= \mathbb{D} + \frac{1}{2}\mathbb{D}\nonumber\\
    &= \frac{3}{2}\mathbb{D}
\end{align}
In the following, we will forgo the delay values and present results with respect to the adjusted \glspl{TTF}.

@@

\medskip
The \gls{WCA} task employed for this study corresponded to a modified and extended version of the LEGO assembly task introduced in~\cite{chen2015early}.
Unlike the original design of this task, in which the user is led through the assembly of a specific model, the modified LEGO task consisted of a pseudo-random sequence of steps instructing subjects to add or remove pieces.

We introduced an experimental design component denoted a \emph{block}, illustrated in \cref{fig:stepblock}, corresponding to a segment of consecutive steps subject to the same delay value.
To study the effects of a delay applied across multiple steps, we manipulated the length of blocks during the task. %, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
Furthermore, we defined a \emph{slice} as a continuous segment of \num{4} steps within a block;
\num{4}-step blocks were composed of a single slice, whereas \num{8}- and \num{12}-step blocks were composed respectively of \num{2} and \num{3} consecutive slices.
Using these experimental design components, we generated a task for each participant by arranging blocks according to a \emph{Latin-square}~\cite{keedwell2015latin} design.
%We generated a pseudo-random permutation of the combinations of block length and delay values, and assigned a unique sequence of steps to each combination in order to create \num{21} unique blocks of steps.
%Using a \emph{Latin-square}~\cite{keedwell2015latin} design, we reordered the initial permutation to generate \num{40} tasks.

To measure the extent to which individual differences in each subject affected obtained task-performance results, participants were asked to complete two questionnaires previous to beginning the task;
the \gls{BFI}~\cite{john1999big} and the \gls{ITQ}~\cite{witmer1998measuring}.
These correspond to well-established tools for the measurement of individual personality traits in humans.
%The \gls{BFI}~\cite{john1999big} consists of \num{44} questions assessing the traits of
%\begin{inlineenum}
%    \item agreeableness
%    \item conscientiousness
%    \item extroversion
%    \item neuroticism
%    \item openness
%\end{inlineenum}.
%The \gls{ITQ}~\cite{witmer1998measuring}, on the other hand, assesses involvement, focus, and the tendency to play games, using \num{29} questions.
The scores for these questionnaires were normalized in post-processing to fall in the \ensuremath{[0, 1]} range for ease of interpretation.

In order to try to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
This array included devices to capture a number of physiological measures, including \gls{EEG}, \gls{HR}, and \gls{GSR}.
%\begin{inlineenum}
%    [itemjoin={{, }}, itemjoin*={{, and }}, label={(\arabic*)}]
%    \item \gls{GSR} (also known as electrodermal activity)
%    \item accelerometer data from the dominant wrist
%    \item brain activity in the form of \gls{EEG}
%    \item heart rate
%\end{inlineenum}.
Unfortunately, no statistically significant effects were measured in any of these variables, and as such neither will be discussed in this summary.
Please refer to \cref{paper:olguinmunoz2021impact} for details.

Finally, we collected video recordings of participants' faces and as well as of the live feed of frames provided as input to the \gls{WCA}.
The facial recordings will also not be discussed in this section, as they were not employed in any analyses.
However, the recording of the live feed was employed later on in the implementation of a procedure for the generation of dynamic synthetic traces for the LEGO task in the context of the EdgeDroid load generator for \gls{WCA}.
This will be discussed in \cref{model:traces}.

@@

\Cref{fig:ttfvsexectime} shows effects on the measured execution times of participants as perceived \gls{TTF} increased.
These results have been grouped by block length (\cref{fig:ttfvsexectime:block}), showing the cumulative effects of delay over series of steps, and slice number (\cref{fig:ttfvsexectime:slice}), showing the evolution of these effects within a series of steps at the same level of impairment.

\cref{fig:ttfvsexectime:block} shows that as system responsiveness decreased, the speed at which participants completed instructions tended to slow down.
Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.
\Cref{fig:ttfvsexectime:block} illustrates this effect clearly;
for all block lengths, as \gls{TTF} grows we see an accompanying increase in execution times.
Moreover, this effect seemed stronger for longer blocks at higher levels of impairment, indicating a slow-down which compounds with the number of steps.
Note that by measuring \emph{step execution time} we are \emph{per se} compensating for the added delay in the measure, and thus this trend must result from the participants' behavioral slow-down as a response to the perceived increase in \gls{TTF}.
This was confirmed through an \gls{ANOVA}~\cite{fujikoshi1993two} including factors of block length and \gls{TTF}; significant effects were found for both factors and their interaction (\ensuremath{p < 0.01}).

On the other hand, in \cref{fig:ttfvsexectime:slice} we observe the evolution of execution times over \emph{slices} of \num{4} steps within blocks.
This figure illustrates how, at the lowest levels of impairment, users tended to get faster as the task progressed.
At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block was roughly \ensuremath{\sim\SI{26}{\percent}} lower than for steps \num{1} to \num{4}.
However, this effect is strongly dampened at higher levels of impairment, with the difference between the average execution times of steps in slices \num{2} and \num{3} basically disappearing at the highest \gls{TTF}.
This was also confirmed through an \gls{ANOVA} on factors slice and \gls{TTF}, finding significant (\ensuremath{p < 0.01}) for both factors and their interaction.

\begin{figure}
    \begin{minipage}[t]{.45\textwidth}
        \centering
        \includegraphics[height=13em]{Figs/2021Impact/previousblock_vs_exectime}
        \caption{Mean execution time of the first slice of steps in a block, grouped by \acs{TTF} of the block immediately preceding the current one.}\label{fig:prevblockvsexectime}
    \end{minipage}%
    \hfill%
    \begin{minipage}[t]{.45\textwidth}
        \centering
        \includegraphics[height=13em]{publications/2023EdgeDroid2/figs/new_model/mu_fits_exgaussian_slice0}
        \caption{%
            Illustration of the effect of neuroticism on the \ensuremath{\mu} parameter of \acs{exGaussian} distributions fitted to the execution times of steps in slice \num{1}.
            Fits were achieved through \gls{MLE}.
            Figure originally published in \cref{paper:olguinmunoz2023realistic}
        }\label{fig:ttfvsexgaussianmu}
    \end{minipage}
\end{figure}

As mentioned before, we found that the effects of system responsiveness on execution times lingered on after the system moved to a different level of impairment.
\Cref{fig:prevblockvsexectime} illustrates this effect.
On average, the mean execution time of the first four steps of blocks was roughly \SI{7}{\percent} lower when the previous block was at a \ensuremath{TTF \leq \SI{2.475}{\second}} compared to when the previous block was at a \ensuremath{TTF > \SI{2.475}{\second}} (single-tailed \ensuremath{p < 0.001}).

Analyses of the correlation between individual subject differences and the above effects highlighted the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
This trait has been linked to low tolerance for stress, high emotional reactivity, as well as higher rates of \emph{delay discounting}~\cite{hirsh2008delay}.
Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \ensuremath{\sim\SI{73}{\percent}} of the variance in the execution time results, neuroticism was present in the first two.
Individual neuroticism scores were found to significantly linearly correlate (\ensuremath{\rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.

>>>>

In this section, I aim to delve deeper into the implications of the methodology for the optimization of \gls{WCA}.
In particular, I wish to explore the impact improving the accuracy of the client-side emulation has on the potential for optimization of the applications under evaluation.
I approach this question by developing an improved, more detailed version of the model of human behavior in \gls{WCA} used in the EdgeDroid tool, which I then incorporate into my methodology along with a mathematical framework to study the potential for optimizing \gls{WCA} applications.

The work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} has used a model that emulates a rigid and unyielding human who makes no mistakes and is not affected by low responsiveness in the system.
However, this is not an accurate representation of real human behavior, and a more realistic human model must take into consideration the perceived feedback latency and its effects on human reactions.
In order to develop a more accurate human model in the context of \gls{WCA}, I first conduct an experimental study that measures human performance in different scenarios, taking into account the impact of system latency on the behavior of humans.
The results of this study serve as the foundation for building a more representative model of human behavior.

Once I have established a more accurate model of human behavior, I can explore its implications for the optimization of \gls{WCA} through my methodology.
By improving the accuracy of the model of human behavior, I aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how the methodology can be effectively applied to address these.
This research highlights the importance of the accuracy of emulations as  crucial factor in the application of my methodology.

@@

In order to more easily and accurately discuss the implications of my methodology for \gls{WCA}, I first characterize these applications and provide some definitions relating to their operation.
I focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistive applications which have as their goal the guidance of a user through a task, using a series of instructions.

As discussed above, \glspl{WCA} track the state of the real world through sampling of inputs, most commonly of video feeds.
In step-based \glspl{WCA}, the state in question refers to the progress of some arbitrary, sequential task, such as the assembly of a LEGO model or of a piece of IKEA furniture~\cite{chen2018application}.
I will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided;
the time interval between these two events will be referred to as the \emph{step duration}.

@@

In the following, please refer to \cref{fig:wcastep}.
\ensuremath{\{ t_0, t_1, \ldots, t_{n + 1} \}} corresponds to a series of discrete and sequential sampling instants, such that \ensuremath{t_0} corresponds to the instant at which the instruction for step \ensuremath{M} is provided and the first sample of the step is taken.
\ensuremath{t_n} corresponds to the instant at which the final sample of step \ensuremath{M} is taken;
this sample is the first sample to capture the finished state of the step.
At this point, the application transitions to the next step \ensuremath{M + 1}, and thus \ensuremath{t_{n + 1}} is the instant at which the next instruction is provided and step \ensuremath{M + 1} begins.
I will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.

\glsreset{TTF}%
From this characterization, key variables relating to human behavior with respect to the sampling behavior of these applications can be defined.
However, it must first be noted that the combination of the characteristic of seamless interaction with the context of the user described in \cref{sec:background:wca} and in \cref{fig:wca:operation}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
A corollary of this is that these are therefore also the \emph{only} points in time at which users can notice changes in system responsiveness.
If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
\begin{inlineenum}
    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
    \item the interval \ensuremath{t_{n + 1} - t_c} I will term \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
\end{inlineenum}
These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
I will refer to them extensively in \cref{ssec:characterizing} onwards.

\subsection{Characterizing human behavior}\label{ssec:characterizing}

\Cref{paper:olguinmunoz2021impact} aims to investigate the effects of system responsiveness on human behavior in step-based \gls{WCA}.
This corresponds to the first step in extending the EdgeDroid tool for the benchmarking of \gls{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.
The study involves \num{40} undergraduate students who interact with a \gls{WCA} application while system responsiveness is systematically altered in real-time, and their behavioral and physiological reactions are recorded.
The study attempts to answer a number of core research questions relating to human behavior in the context of these applications.
These include whether
\begin{inlineenum}
    \item users change their actions in relation to system latency
    \item they show signals of arousal in physiological responses
    \item their responses to delay effects are mediated by cognition and/or emotion
    \item these effects are mediated by personality indicators
\end{inlineenum}.

I find that reduced system responsiveness induces additional behavioral slow-down in \gls{WCA} users, and that this effect scales with the strength of the reduction in system responsiveness.
Furthermore, humans tend to speed up as the task progresses;
however, as system responsiveness decreases, this effect is strongly dampened.
The effects of reduced system responsiveness on users linger for at least four steps after the system responsiveness improves, leading to longer execution times for steps immediately after a period of reduced responsiveness.
Finally, the personality trait of \emph{neuroticism}~\cite{john1999big} seems to play a significant role in modulating users' response to reduced responsiveness in \gls{WCA}.

@@

I employ a version of the step-based \emph{Gabriel} cognitive assistant framework first introduced by \citeauthor{chen2018application}~\cite{chen2018application}.
The assistant is instrumented to capture key application and task performance metrics in real time, in particular step execution time.
Furthermore, the software is modified to allow for the real-time alteration of system responsiveness by extending the processing interval of each input to a temporarily fixed value denoted \emph{delay}.
This is illustrated in \cref{fig:regularwca-vs-delaywca}.
If, during a series of steps, delay is set to a value \ensuremath{\mathbb{D}}, the feedback for each input frame is provided to the user exactly a time interval \ensuremath{\mathbb{D}} after frame capture.
That is,
\begin{align}
    t_k - t_{k - 1} = \mathbb{D} \qquad \forall k \in [2, n + 1]
\end{align}
%Seven values were selected for \ensuremath{\mathbb{D}}: no added delay (which was referred to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second}.
In order to fully characterize the resulting \emph{\glspl{TTF}} perceived by participants, we note the instant of final frame capture for a step, \ensuremath{t_c} (and conversely the wait time \( \mathcal{W} \) of the step), can be assumed to be uniformly distributed in \( [t_{n - 1}, t_n] \), without loss of generality.
Thus the average \gls{TTF} for a step can be expressed as
\begin{align*}
    TTF = (t_{n + 1} - t_{n}) + \frac{1}{2}(t_n - t_{n - 1})
\end{align*}
which, for a step with a given delay \ensuremath{\mathbb{D}}, works out to\footnote{%
    The \num{0}/no-delay case was handled by calculating the empirical average frame \gls{RTT} for these steps, and using this value as the delay \ensuremath{\mathbb{D}}.
    This yielded an average \gls{TTF} of \SI{0.57}{\second} for these steps.
}
\begin{align}
    TTF &= \mathbb{D} + \frac{1}{2}\mathbb{D}\nonumber\\
    &= \frac{3}{2}\mathbb{D}
\end{align}
In the following, I forgo the delay values and present results with respect to the adjusted \glspl{TTF}.

@@

\medskip
The \gls{WCA} task employed for this study corresponds to a modified and extended version of the LEGO assembly task introduced in~\cite{chen2015early}.
Unlike the original design of this task, in which the user is led through the assembly of a specific model, the modified LEGO task consists of a pseudo-random sequence of steps instructing subjects to add or remove pieces.

I introduce an experimental design component denoted a \emph{block}, illustrated in \cref{fig:stepblock}, corresponding to a segment of consecutive steps subject to the same delay value.
To study the effects of a delay applied across multiple steps, I manipulate the length of blocks during the task. %, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
Furthermore, I define a \emph{slice} as a continuous segment of \num{4} steps within a block;
\num{4}-step blocks are composed of a single slice, whereas \num{8}- and \num{12}-step blocks are composed respectively of \num{2} and \num{3} consecutive slices.
Using these experimental design components, I generate a task for each participant by arranging blocks according to a \emph{Latin-square}~\cite{keedwell2015latin} design.

To measure the extent to which individual differences in each subject affect obtained task-performance results, participants are asked to complete two questionnaires previous to beginning the task;
the \gls{BFI}~\cite{john1999big} and the \gls{ITQ}~\cite{witmer1998measuring}.
These correspond to well-established tools for the measurement of individual personality traits in humans.
%The \gls{BFI}~\cite{john1999big} consists of \num{44} questions assessing the traits of
%\begin{inlineenum}
%    \item agreeableness
%    \item conscientiousness
%    \item extroversion
%    \item neuroticism
%    \item openness
%\end{inlineenum}.
%The \gls{ITQ}~\cite{witmer1998measuring}, on the other hand, assesses involvement, focus, and the tendency to play games, using \num{29} questions.
The scores for these questionnaires are normalized in post-processing to fall in the \ensuremath{[0, 1]} range for ease of interpretation.

In order to try to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants are asked to wear an array of biometric sensors.
This array includes devices to capture a number of physiological measures, such as \gls{EEG}, \gls{HR}, and \gls{GSR}.
%\begin{inlineenum}
%    [itemjoin={{, }}, itemjoin*={{, and }}, label={(\arabic*)}]
%    \item \gls{GSR} (also known as electrodermal activity)
%    \item accelerometer data from the dominant wrist
%    \item brain activity in the form of \gls{EEG}
%    \item heart rate
%\end{inlineenum}.
Unfortunately, no statistically significant effects are measured in any of these variables, and as such neither is discussed in this summary.
Please refer to \cref{paper:olguinmunoz2021impact} for details.

Finally, I collect video recordings of participants' faces and as well as of the live feed of frames provided as input to the \gls{WCA}.
The facial recordings are also not discussed in this section, as they are not employed in any analyses.
However, the recording of the live feed is discussed in \cref{model:traces}, in the context of an implementation of a procedure for the generation of dynamic synthetic traces for the LEGO task in the context of the EdgeDroid load generator for \gls{WCA}.

@@

\Cref{fig:ttfvsexectime} shows effects on the measured execution times of participants as perceived \gls{TTF} increases.
These results have been grouped by block length (\cref{fig:ttfvsexectime:block}), showing the cumulative effects of delay over series of steps, and slice number (\cref{fig:ttfvsexectime:slice}), showing the evolution of these effects within a series of steps at the same level of impairment.

\cref{fig:ttfvsexectime:block} shows that as system responsiveness decreases, the speed at which participants complete instructions tends to slow down.
Compared to the unimpaired case, participants are on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.
\Cref{fig:ttfvsexectime:block} illustrates this effect clearly;
for all block lengths, as \gls{TTF} grows an accompanying increase in execution times can be observed.
Moreover, this effect seems stronger for longer blocks at higher levels of impairment, indicating a slow-down which compounds with the number of steps.
Note that by measuring \emph{step execution time} we are \emph{per se} compensating for the added delay in the measure, and thus this trend must result from the participants' behavioral slow-down as a response to the perceived increase in \gls{TTF}.
This is confirmed through an \gls{ANOVA}~\cite{fujikoshi1993two} including factors of block length and \gls{TTF}; significant effects are found for both factors and their interaction (\ensuremath{p < 0.01}).

On the other hand, in \cref{fig:ttfvsexectime:slice} the evolution of execution times over \emph{slices} of \num{4} steps within blocks is illustrated.
This figure illustrates how, at the lowest levels of impairment, users tend to get faster as the task progresses.
At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block is roughly \ensuremath{\sim\SI{26}{\percent}} lower than for steps \num{1} to \num{4}.
However, this effect is strongly dampened at higher levels of impairment, with the difference between the average execution times of steps in slices \num{2} and \num{3} basically disappearing at the highest \gls{TTF}.
This is also confirmed through an \gls{ANOVA} on factors slice and \gls{TTF}, finding significant (\ensuremath{p < 0.01}) for both factors and their interaction.

\begin{figure}
    \begin{minipage}[t]{.45\textwidth}
        \centering
        \includegraphics[height=13em]{Figs/2021Impact/previousblock_vs_exectime}
        \caption{Mean execution time of the first slice of steps in a block, grouped by \acs{TTF} of the block immediately preceding the current one.}\label{fig:prevblockvsexectime}
    \end{minipage}%
    \hfill%
    \begin{minipage}[t]{.45\textwidth}
        \centering
        \includegraphics[height=13em]{publications/2023EdgeDroid2/figs/new_model/mu_fits_exgaussian_slice0}
        \caption{%
            Illustration of the effect of neuroticism on the \ensuremath{\mu} parameter of \acs{exGaussian} distributions fitted to the execution times of steps in slice \num{1}.
            Fits are achieved through \gls{MLE}.
            Figure originally published in \cref{paper:olguinmunoz2023realistic}
        }\label{fig:ttfvsexgaussianmu}
    \end{minipage}
\end{figure}

As mentioned before, the effects of system responsiveness on execution times linger on after the system moves to a different level of impairment.
\Cref{fig:prevblockvsexectime} illustrates this effect.
On average, the mean execution time of the first four steps of blocks is roughly \SI{7}{\percent} lower when the preceding block was at a \ensuremath{TTF \leq \SI{2.475}{\second}} compared to when the preceding block was at a \ensuremath{TTF > \SI{2.475}{\second}} (single-tailed \ensuremath{p < 0.001}).

Analyses of the correlation between individual subject differences and the above effects highlight the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
This trait has been linked to low tolerance for stress, high emotional reactivity, as well as higher rates of \emph{delay discounting}~\cite{hirsh2008delay}.
Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \ensuremath{\sim\SI{73}{\percent}} of the variance in the execution time results, neuroticism is present in the first two.
Individual neuroticism scores are found to significantly linearly correlate (\ensuremath{\rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
